---
title: "Initial data cleanup (WIP)"
author: "Virginia Scarlett"
date: "5/23/2025"
---

## Overview
This script reads in the raw survey data from Qualtrics and splits the data into multiple files. It assumes the data were exported from Qualtrics using 'More Options' > 'Split multi-value fields into columns'.\
\
It also assumes the data file is called 'raw_survey_data.tsv'.\
\
I'm storing the path to my data folder in .Renviron like so: DATA_PATH = "/Path/to/data/folder"\
Because the data contain personally identifiable information, I'm grabbing them from elsewhere in my filesystem, outside of this project directory.

## Import packages and utilities
```{r}
project_root <- here::here() # requires that you be in project directory if using VS Code (i.e. not home dir)
print(paste0("Project root: ",project_root))
suppressMessages(source(file.path(project_root, "scripts/packages.R")))
suppressMessages(source(file.path(project_root, "scripts/utils.R")))
```

## Define functions specific to this script
```{r}
write_subset_of_data <- function(df, filen) {
  write.table(df,
    file.path(Sys.getenv("DATA_PATH"), filen),
    quote = FALSE,
    row.names = FALSE,
    sep = "\t"
  )
}
```

## Load data

Open the raw survey data from Qualtrics. N.B. Qualtrics exports in UTF-16.
```{r}
data <- load_qualtrics_data("raw_survey_data.tsv", fileEncoding = "utf-16")
```

## Inspect data
As mentioned above in [Overview], we used Qualtrics' 'Split multi-value fields into columns' function when we exported the data from Qualtrics, which causes every response field to be its own column. So if a multiple-choice question has 4 possible responses, it will have 4 columns in the dataframe.\
Here are the dimensions of the data:
```{r}
dim(data)
```
Here are the columns, as a vector:
```{r}
names(data)
```
The column names are based on the question names in Qualtrics (which I chose). I don't know why the numbers at the end are kind of arbitrary for certain questions. I'm pretty sure Qualtrics automatically added that "_TEXT" suffix to the free-response question columns.
