To address a peer review comment, I am comparing the number of respondents per campus to the campus population. Data from the UC website is hard-coded into here.ß

# Import packages and utilities
```{r}
project_root <- here::here() # requires that you be somewhere in the
# project directory (not above it)
# packages
suppressMessages(source(file.path(project_root, "scripts/packages.R")))
# functions and objects used across scripts
suppressMessages(source(file.path(project_root, "scripts/utils.R")))
```

## Load data

```{r}
other_quant <- load_qualtrics_data("clean_data/other_quant.tsv")
status <- load_qualtrics_data("clean_data/contributor_status_Q3.tsv")

raw_data <- cbind(status, other_quant)
nrow(raw_data)
head(raw_data)
```

## Wrangle data

Filter out people who were neither past nor future contributors. We'll use this filtered data frame moving forward.
```{r}
# Filter duds
data <- raw_data %>%
    filter(!(Past == "" | Future == "")) %>%
    filter(!(Past == "False" & Future == "False"))

nrow(data)
```

Count how many responses each campus received.

```{r}
campus_counts <- data %>%
  count(campus, name = "respondents")
```

## Get UC data into data frames
First, let's get the number of students per campus from this website: https://www.universityofcalifornia.edu/about-us/information-center/fall-enrollment-glance
```{r}
campus <- c(
  "UC Berkeley",
  "UC Davis",
  "UC Irvine",
  "UC Los Angeles",
  "UC Merced",
  "UC Riverside",
  "UC San Diego",
  "UC San Francisco",
  "UC Santa Barbara",
  "UC Santa Cruz",
  "Other UC"
)
numStudents <- c(
  46151,
  39409,
  37268,
  47435,
  8943,
  27633,
  45087,
  2993,
  26034,
  20140,
  0
)
student_pops <- data.frame(
  campus,
  numStudents
)
```
Great, this comes out to 301,093, the total reported in this news article: https://www.universityofcalifornia.edu/press-room/university-california-sets-enrollment-records.

Next, let's get the employees. I'll use this table. https://www.universityofcalifornia.edu/about-us/information-center/uc-employee-headcount. This will be imperfect because who knows how many students are also employees. I am discarding the "student teachers/research assistant" category, since that group may be predominantly students (a definition is not provided) who would primarily identify as students when taking our survey.
```{r}
numAcademicStaff <- c(
  5134, #9765-4631
  5951, #9264-3313
  4887, #7533-2646
  9425, #13164-3739
  648, #1405-757
  1867, #3864-1997
  6632, #10540-3908
  8389, #8935-546
  1907, #4109-2202
  1516, #999-1483
  385 #UCOP+ANR
)
acad_staff_pops <- data.frame(
  campus,
  numAcademicStaff
)
```

Likewise, counting up the non-academic staff, I am excluding the "student staff" category.
```{r}
numNonacademicStaff <- c(
  9116, #14557-5441
  23364, #29013-5649
  19991, #23115-3124
  38916, #44446-5530 ASUCLA+UCLA
  1407, #2754-1347
  3324, #5802-2478
  25792, #29862-4070
  29585, #29792-207
  3631, #7086-3455
  2998, #5350-2352
  3026 #3053-27 UCOP+ANR
)
non_acad_staff_pops <- data.frame(
  campus,
  numNonacademicStaff
)
```

## Get UC data into one data frame

Now get the total population of each campus (roughly estimated).
```{r}
merged <- merge(
  merge(student_pops, acad_staff_pops, by = "campus", all = TRUE),
  non_acad_staff_pops,
  by = "campus",
  all = TRUE
)

merged$total <- rowSums(merged[, -1], na.rm = TRUE) #ignore campus column

merged
```

```{r}
total_pops <- merged[,-c(2,3,4)]
```

## Combine survey data with the UC workforce/enrollment data

```{r}
final_data <- merge(campus_counts, total_pops, by = "campus", all = TRUE)
final_data
```

We had no respondents from UC Riverside. Let's change that NA to 0.

```{r}
final_data[is.na(final_data)] <- 0
```

Sort and reorder factor levels

```{r}
final_data <- final_data[order(-final_data$total), ]
final_data$campus <- factor(final_data$campus, levels=final_data$campus)
final_data
```

Let's create a trend line and show the R squared.

```{r}
fit <- lm(respondents ~ total, data = final_data)
r_sq <- summary(fit)$r.squared

# Format the text so it looks nice in the plot
# Using "==" tells ggplot to format it as a proper math equation
r_label <- sprintf("R^2 == %.3f", r_sq)
```

By default, the data are very spread out across the plot so it's hard to take in the whole dataset. I'll zoom out to make the dataset easier to see. Here I'll adjust the amount of padding I want to add to the x and y axes.

```{r}
pad_prop <- 0.15 

# Calculate the maximums and add the padding proportion
max_x <- max(final_data$total) * (1 + pad_prop)
max_y <- max(final_data$respondents) * (1 + pad_prop)
```

```{r fig.width=9, fig.height=8}
final_plot <- ggplot(final_data, aes(x = total, y = respondents)) +
  geom_point(color = "#651FFF", size = 2) +
  scale_x_continuous(labels = label_comma()) + # Add commas to x-axis
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.5) +

  annotate(
    "text",
    x = max_x,
    y = max_y,
    label = r_label,
    parse = TRUE, # Turns "R^2 ==" into actual math formatting
    hjust = 1, # Right-aligns the text
    vjust = 1, # Top-aligns the text
    size = 5
  ) +

  # Add campus labels
  ggrepel::geom_text_repel(
    aes(label = campus),
    size = 2.4,
    box.padding = 0.5,
    point.padding = 0.5,
    segment.color = "grey50"
  ) +

  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text.x = element_text(angle = 60, vjust = 0.6, size = 6),
    axis.text.y = element_text(size = 6),
    axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(),
    legend.title = element_blank(),
    plot.title = element_text(
      hjust = 0,
      face = "bold",
      size = 11,
      margin = margin(b = 5)
    ),
    plot.margin = unit(c(0.3, 0.3, 0.3, 0.9), "cm"),
    panel.grid = element_line(linetype = "solid", color = "gray90"),
    panel.background = element_blank()
  ) +

  labs(
    title = "Number of respondents vs. campus population size",
    x = "Population size (staff + students)",
    y = "Number of survey respondents"
  ) +
  # Pass the calculated limits into coord_cartesian
  coord_cartesian(xlim = c(0, max_x), ylim = c(0, max_y))

  final_plot
```

```{r}
ggsave(
  filename = file.path(FIGURE_PATH, "response_campus_corr.tif"),
  plot = final_plot,
  device = "tiff",
  width = 5.2, height = 5, units = "in",
  dpi = 310,                
  compression = "none",    
  bg = "white"
)
```
