---
title: "Challenges + job category"
---

# Overview

Secondary analysis of survey Q9: "How frequently have you encountered the following challenges while working on open-source projects?"

In this script, I am considering challenges in light of job category.

## Import packages and utilities

```{r}
project_root <- here::here() # requires that you be somewhere in the
# project directory (not above it)
# packages
suppressMessages(source(file.path(project_root, "scripts/packages.R")))
# functions and objects used across scripts
suppressMessages(source(file.path(project_root, "scripts/utils.R")))
```


## Load data
```{r}
challenges <- load_qualtrics_data("clean_data/challenges_Q9.tsv")
other_quant <- load_qualtrics_data("clean_data/other_quant.tsv")
```

## Wrangle data
```{r}
challenges_and_job <- challenges
challenges_and_job$job_category <- other_quant$job_category

head(challenges_and_job)
```

Remove rows that contain any empty entries.
```{r}
nrow(challenges_and_job)
challenges_and_job <- exclude_empty_rows(challenges_and_job, strict = TRUE) # from scripts/utils.R
nrow(challenges_and_job)
```


For visual clarity in our plots, let's combine postdocs and other staff researchers.
```{r}
challenges_and_job <- challenges_and_job %>%
  mutate(
    job_category = recode(
      job_category,
      "Post-Doc" = "Postdocs and\nStaff Researchers",
      "Other research staff" = "Postdocs and\nStaff Researchers"
    )
  )

head(challenges_and_job)
```

Let's reshape the data from wide to long format for easier counting and plotting.
```{r}
long_data <- challenges_and_job %>%
  pivot_longer(
    cols = -last_col(),
    names_to = "challenge",
    values_to = "challenge_level"
  )

long_data
```

Since it's overwhelming to look at the distribution of challenge levels for all groups, let's just look at the proportion of that group who said "frequently" or "always".
```{r}
# Calculate proportion of TRUEs by taking the mean of a logical vector,
# created by %in%.
to_plot <- long_data %>%
  group_by(job_category, challenge) %>%
  summarize(proportion = mean(challenge_level %in% c("Frequently", "Always"))) %>%
  ungroup()
to_plot
```

Calculate the standard deviation for each challenge and reorder the factor levels by stdev in our plot. (It looks nicer.)
```{r}
stdev_df <- to_plot %>%
  group_by(challenge) %>%
  summarise(
    st_dev = sd(proportion, na.rm = TRUE)
  ) %>%
  ungroup()

# Order by stdev
stdev_df <- stdev_df %>%
    arrange(desc(st_dev))

# Reorder factor levels
to_plot$challenge <- factor(to_plot$challenge, levels = stdev_df$challenge)
```
```{r, fig.width=9, fig.height=6}
detailed_challenges_plot <- ggplot(to_plot, aes(x = challenge, y = proportion, group = job_category, color = job_category, shape = job_category)) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = "Challenge",
    y = "Proportion saying 'Frequently' or 'Always'",
    color = "Job Category",
    shape = "Job Category",
    title = "Proportion of respondents frequently facing challenges"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
detailed_challenges_plot
```

Save, if you wish.
```{r}
#save_plot("detailed_challenges_by_job.tiff", 12, 10, p=detailed_challenges_plot)
```

That's a nice plot, but it's probably too information-dense for a presentation, or even a paper. Let's just look at the top 3 challenges for each group.
```{r}
top3 <- to_plot %>%
  group_by(job_category) %>%
  slice_max(order_by = proportion, n = 3)
```

```{r}
# Filter to include only challenges present in the top3 dataframe 
filtered_plot <- to_plot %>%
  semi_join(top3, by = c("job_category", "challenge"))
```

```{r}
# Reorder fill factor levels so legend items are in order of appearance
desired_levels <- top3 %>% 
  pull(challenge) %>% 
  unique()

filtered_plot <- filtered_plot %>%
  mutate(
    challenge = factor(challenge, levels = desired_levels)
  )
```

```{r}
# Reorder x-axis factor levels to match academic advancement
job_level_order <- c(
  "Faculty",
  "Postdocs and\nStaff Researchers",
  "Grad Student",
  "Undergraduate",
  "Non-research Staff"
)
filtered_plot$job_category <- factor(
  filtered_plot$job_category, 
  levels = job_level_order
  )
```


```{r, fig.width=9, fig.height=6}
job_challenge_plot <- ggplot(
  filtered_plot,
  aes(
    x = job_category,
    y = proportion,
    fill = challenge
  )
) +
  geom_col(position = position_dodge()) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_fill_manual(values = COLORS) +
  labs(
    x = "Job Category",
    y = "Proportion saying\n'Frequently' or 'Always'",
    fill = "Challenge",
    title = "Top 3 Challenges by Job Category"
  ) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 24),
    axis.text.x = element_text(angle = 60, vjust = 0.6, size = 18),
    axis.text.y = element_text(size = 18),
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    panel.background = element_blank(),
    panel.grid = element_line(linetype = "solid", color = "gray90"),
    plot.title = element_text(hjust = 0.5, size = 24),
    plot.margin = unit(c(0.3, 0.3, 0.3, 0.3), "cm")
  )
job_challenge_plot
```

Save, if you wish.
```{r}
#save_plot("top3_challenges_by_job.tiff", 12, 10, p=job_challenge_plot)
```