---
title: "Demographics: descriptive statistics"
---

# Overview

Let's look at survey participation rates across various groups (demographics). This isn't about hypothesis testing; it's just about getting a sense of who took the survey.

# Import packages and utilities
```{r}
project_root <- here::here() # requires that you be somewhere in the
# project directory (not above it)
# packages
suppressMessages(source(file.path(project_root, "scripts/packages.R")))
# functions and objects used across scripts
suppressMessages(source(file.path(project_root, "scripts/utils.R")))
```

# Load data

```{r}
other_quant <- load_qualtrics_data("clean_data/other_quant.tsv")
status <- load_qualtrics_data("clean_data/contributor_status_Q3.tsv")
qual <- load_qualtrics_data("qual_responses.tsv")

data <- cbind(status, other_quant)
nrow(data)
head(data)
```

# Experienced vs aspiring

First, let's see how many experienced and aspiring open source contributors took the survey.

Experienced:
```{r}
nrow(subset(data, Past=="True"))
```

Aspiring:
```{r}
nrow(subset(data, Past=="False" & Future=="True"))
```

# Campus

I already learned while plotting the data that UCSB and UCLA are overrepresented. What proportion of respondents came from these two schools?
```{r}
unique(data$campus)
```

Let's remove people who said "I'm not affiliated with UC".

```{r}
data <- subset(data, campus != "I'm not affiliated with UC")
nrow(data)
```

Only 2 respondents were not UC affiliates.
```{r}
campus_count <- data.frame(table(data$campus))
names(campus_count) <- c("Campus", "Count")
total <- sum(campus_count$Count)
ucsb <- subset(campus_count, Campus=="UC Santa Barbara")[,"Count"]
ucla <- subset(campus_count, Campus=="UC Los Angeles")[,"Count"]
round((ucsb + ucla) / total * 100, digits = 1)
```

So 48% of respondents came from these two campuses.

# Field of study
How many respondents were from STEM, social science, and humanities?

```{r}
# Remove people who didn't answer this question--non-research staff
tmp <- data$field_of_study[nzchar(data$field_of_study)]
field_count <- data.frame(table(tmp))
names(field_count) <- c("Field", "Count")
total <- sum(field_count$Count)
stem <- sum(
  subset(
    field_count,
    Field == "Life sciences" |
    Field == "Math and CS" |
    Field == "Physical sciences"
  )[, "Count"]
)
sosc_hum <- sum(
  subset(
    field_count,
    Field == "Humanities" |
    Field == "Social sciences"
  )[, "Count"]
)
# sanity check
total == stem + sosc_hum

round(stem / total * 100, digits = 1)
round(sosc_hum / total * 100, digits = 1)
```

So 86% of respondents are from stem, and 14% are from social sciences/humanities.

How many of the STEM respondents are from math or CS?

```{r}
math_cs <- sum(
  subset(
    field_count,
    Field == "Math and CS"
  )[, "Count"]
)
round(math_cs / stem * 100, digits = 1)
```

53% of STEM respondents are from math or CS.

How many experienced contributors were from humanities or social sciences?

```{r}
nrow(subset(data, field_of_study=="Humanities" & Past =="True"))

nrow(subset(data, field_of_study=="Social sciences" & Past =="True"))
```

We had 4 experienced contributors from the humanities, and 10 from the social sciences. 

# Job category

```{r}
# Remove people who didn't answer this question--
# neither future nor past contributors
tmp <- data$job_category[nzchar(data$job_category)]
job_count <- data.frame(table(tmp))
names(job_count) <- c("Job", "Count")
total <- sum(job_count$Count)

nr_staff <- subset(job_count, Job == "Non-research Staff")[, "Count"]
academics <- sum(subset(job_count, Job != "Non-research Staff")[, "Count"])

job_count

round(nr_staff / total * 100, digits = 1)
round(academics / total * 100, digits = 1)
```

36% of survey respondents are non-research staff, while 64% are academics.

What about the job areas of the non-research staff?

```{r}
# Remove everybody except non-research staff
tmp <- data$staff_categories[nzchar(data$staff_categories)]
staff_count <- data.frame(table(tmp))
names(staff_count) <- c("Area", "Count")
total <- sum(staff_count$Count)

staff_count

rs <- subset(staff_count, Area == "Academic and Research Support")[, "Count"]
it <- subset(staff_count, Area == "Information Technology (IT)")[, "Count"]

round( (rs + it) / total * 100, digits = 1)
```

67% of the non-research staff respondents were from either IT or Academic and Research Support, which we told participants "includes research administration, libraries, and instructional design".

# Qualitative responses: staff categories
Let's look at the free-response field for staff job categories. These are the non-research staff who selected "other" and wrote in their job area.

```{r}
qual_staff <- qual$staff_categories_13_TEXT[nzchar(qual$staff_categories_13_TEXT)]
```

I looked at these manually, but for the sake of data privacy, I am not printing the free responses here. I can see that the word "Library" and the abbreviation "IT" each occur multiple times. 

```{r}
length(qual_staff)
sum(str_count(qual_staff, pattern = "Library"))
sum(str_count(qual_staff, pattern = "IT"))
```

So, seven of the free-text responses contained either the word "Library" or "IT" or both. I am looking manually, and I can see that these came from 6 people. (One person put "Library IT".)

# Qualitative responses: subfields

For Q18, after asking them their broad domain of study, we asked them this free-response question:
What is your primary field? (e.g. astrophysics, neuroscience)
One response is preferred, but if multiple are needed, please separate with commas.

```{r}
qual_fields <- qual$subfield[nzchar(qual$subfield)]
length(qual_fields)
academics
```

This question was not mandatory, but 170/188 academics answered it.

```{r}

```