---
title: "Demographics: qualitative responses"
---

At this point, I've manually reviewed the classifications of people's write-in responses to Q18 in Microsoft Excel. There were definitely some decisions that I think even state-of-the-art AI would have struggled with.

I saved my work in an excel spreadsheet. The key column is the "decision" column, so I just copied that one column to a code-friendly text editor and saved it as a tsv. Here's my curation schema:
k = keep as-is
d = drop
If I put a number in the decision column, that means replace the existing row with that row of the taxonomy.
Also, the taxonomy didn't have "AI" or "Machine Learning" by themselves. So I tagged people who wrote in "AI" or "ML" as 1017, "Artificial Intelligence and Robotics", even though none of these people mentioned robotics. I'll correct that right before I plot the data.

We'll want to use this column to make substitutions in the data frame produced by demographics_qual_part1.qmd.

I also recorded 4 rows that need to be added. These were cases where people indicated two fields, usually with an "and" instead of with a slash or comma, so I didn't split these on my first pass. Since there's only 4 of these, I'm just hard-coding them in here. (Ugly code, yeah, I know.)

```{r}
additions <- data.frame(
    participantID = c(16, 62, 137, 159),
    decision = c(1029, 1016, 1093, 838)
)
```

# Import packages and utilities
```{r}
project_root <- here::here() # requires that you be somewhere in the
# project directory (not above it)
# packages
suppressMessages(source(file.path(project_root, "scripts/packages.R")))
# functions and objects used across scripts
suppressMessages(source(file.path(project_root, "scripts/utils.R")))
```


# Load data
```{r}
tax <- as.data.frame(
    readLines("data/digital_commons_disciplines.txt"),
    stringsAsFactors = FALSE)

# Data cleaning
tax <- tax %>%
  separate(
    col   = names(tax)[1],
    into  = c("Level1", "Level2", "Level3"),
    sep   = ": ",
    fill  = "right",   # any missing pieces become NA
    extra = "merge"    # if there were >2 colons, they’d all merge into Level3
  )
```

```{r}
auto_results <- read.csv("data/qual_fields_guesses.tsv", sep="\t", stringsAsFactors = FALSE)

decisions <- as.data.frame(
    readLines("data/classification_decisions.tsv"),
    stringsAsFactors = FALSE)

names(decisions) <- c("decision")
```

Let's make sure that both data frames have the same number of rows.

```{r}
nrow(auto_results) == nrow(decisions)
```

Now, let's start updating the data frame we got in part 1 with the manual revisions. First, let's drop the rows we can drop.

```{r}
dim(auto_results)
curated <- auto_results
curated <- cbind(curated, decisions)

curated <- curated %>%
    filter(decision != "d")

dim(curated)
head(curated)
```

Next, let's substitute the bad rows with the correct ones.

```{r}
# Try to coerce entries in decision to numeric
dec_num <- suppressWarnings(as.integer(curated$decision))
head(dec_num)

# Logical vector indicating whether decision is a (sensible) number
ok <- !is.na(dec_num) & dec_num >= 1 & dec_num <= nrow(tax)
head(ok)

# overwrite Level columns in curated with those cols
# in the corresponding row from tax
curated[ok, c("Level1", "Level2", "Level3")] <- 
  tax[dec_num[ok], c("Level1", "Level2", "Level3")]
```

Let's inspect the results with a side-by-side comparison (well, vertically speaking.) I could make this print-out prettier if I had time and cared enough.

```{r}
for (i in seq(10)) {
    cat("Before:\n")
    # write.table() prints to the console by default and lets you hide headers
    write.table(subset(auto_results, participantID == i), col.names = FALSE, row.names = FALSE, quote = 2)
    cat("After:\n")
    tmp <- subset(curated, participantID == i)
    write.table(tmp[,-ncol(tmp)], col.names = FALSE, row.names = FALSE, quote = 2)
    cat("\n")
}
```

We see that the errors have been corrected.

We don't need this column anymore.
```{r}
curated <- curated %>% select(-c("decision"))
```

Finally, let's add those additional 4 rows that I hard-coded at the top of this document.

```{r}
# Create a "decision" column that just reflects row number
# so that we can join the decision col in additions with that of tax
tmp_tax <- tax %>% mutate(decision = row_number())

new_rows <- additions %>%
  mutate(decision = suppressWarnings(as.integer(decision))) %>%
  left_join(curated, by = "participantID") %>%
  select(-c("Level1", "Level2", "Level3")) %>% # we're just grabbing the response col from `curated`
  left_join(tmp_tax, by = "decision") %>% # grab the Level cols from tax, at the row specified in decision
  select(participantID, response, Level1, Level2, Level3)

curated <- bind_rows(curated, new_rows)
curated <- curated %>% arrange(participantID)

subset(curated, participantID==16)
```

Ok! Now we finally have the qualitative responses smooshed and zhoozhed into our taxonomy.

# Analysis

First, I'd like to know how many participants identified with each unique option for Level1. So how many people had "Life Sciences" for Level1, how many people had "Arts and Humanities", etc.

```{r}
level1_counts <- curated %>%
  distinct(participantID, Level1) %>% # one row per person per Level1
  count(Level1, name = "n_participants") %>%
  arrange(desc(n_participants))

level1_counts
```

Neat! Now let's do the same for the other two levels.

```{r}
level2_counts <- curated %>%
  distinct(participantID, Level2) %>% # one row per person per Level1
  count(Level2, name = "n_participants") %>%
  arrange(desc(n_participants))

level2_counts
```

Cool! I think Level 2 will be most useful, but let's take a look at level 3.

```{r}
level3_counts <- curated %>%
  distinct(participantID, Level3) %>% # one row per person per Level1
  count(Level3, name = "n_participants") %>%
  arrange(desc(n_participants))

level3_counts
```

Acutally, Level 3 is pretty interesting, too.

Let's do a little curation. First, let's remove those NA rows. These were rows where the participant's input matched at level 1 or 2, but not all three levels, so there was an NA for levels 2 and/or 3.
```{r}
level2_counts <- level2_counts %>%
    filter(!is.na(Level2))
level3_counts <- level3_counts %>%
    filter(!is.na(Level3))
```


And let's change that Artificial Intelligence label I mentioned at the top of this document. 

```{r}
level3_counts$Level3[level3_counts$Level3 == "Artificial Intelligence and Robotics"] <- "AI and Machine Learning"
```

Let's rearrange the columns a bit for readability, and put these all back into one data frame.

```{r}
names(level1_counts)[1] <- "Discipline"
names(level2_counts)[1] <- "Discipline"
names(level3_counts)[1] <- "Discipline"

level1_counts <- cbind("Level 1", level1_counts)
level2_counts <- cbind("Level 2", level2_counts)
level3_counts <- cbind("Level 3", level3_counts)

names(level1_counts)[1] <- "Taxonomy Level"
names(level2_counts)[1] <- "Taxonomy Level"
names(level3_counts)[1] <- "Taxonomy Level"

final_data <- rbind(level1_counts, level2_counts, level3_counts)
```

I'm just going to leave them as a table for now. I'll save them with my figures. Remember from utils.R that I've saved the path to my figures, which are on my local computer, in my .Renviron file.

```{r}
write.table(final_data, file.path(Sys.getenv("FIGURE_PATH"), "qual_disciplines.tsv"), row.names = FALSE, quote = FALSE, sep="\t")
```