---
title: "Demographics: qualitative responses"
---

# Qualitative responses: subfields

For Q18, after asking respondents for their broad domain of study, we asked them this free-response question:\
What is your primary field? (e.g. astrophysics, neuroscience)\
One response is preferred, but if multiple are needed, please separate with commas.\

I plan to "classify" these free responses into standardized categories. The taxonomy of academic disciplines that I'll be using is here:\
https://digitalcommons.elsevier.com/en_US/dc-disciplines-taxonomy

I downloaded the PDF and did a very tiny amount of curation to make this list of academic fields and subfields machine-readable. (I left a txt file with my curation notes in the data/ folder.)

I'll use a fuzzy string matching algorithm to classify the responses--essentially, looking for which item in the taxonomy most closely matches the user's entry. Hopefully, there will be very few entries that aren't a good match for anything in the taxonomy, and I can deal with these stragglers manually.

Note that while I am preforming classification, there's no machine learning here.

Note from the future: this code ended up being pretty ugly, with a lot of nested loops. Maybe I could have written less code if I'd used the `fuzzyjoin` package. Oh well. I went for a rudimentary solution, and it worked well enough.

# Import packages and utilities
```{r}
project_root <- here::here() # requires that you be somewhere in the
# project directory (not above it)
# packages
suppressMessages(source(file.path(project_root, "scripts/packages.R")))
# functions and objects used across scripts
suppressMessages(source(file.path(project_root, "scripts/utils.R")))
```

# Define funtions

#### get_all_hits
- Arguments:
  - `query`: A survey response, e.g. "neuroscience" or "AI".
  - `tax_df`: The taxonomy, as a data frame with columns `Level1`, `Level2`,
    and `Level3`. It wasn't really necessary to include this as an argument
    since it doesn't change, but I like to be explicit about what goes into the
    function.
- Details:
  - For a given query (survey response), return a data frame with the similarity
    scores between that query and all strings in the DC taxonomy. Each row
    in the taxonomy gets a row score, which is the minimum distance between the
    query and any of the three strings in that row. Row score was computed with
    `get_row_results()`.
- Outputs:
  - A data frame with columns `min_dist`, `level`, and `index`. `index` is just
    row number for that row in the taxonomy/in this data frame (they have the
    same number of rows). `level` indicates which level the winning string came
    from. `index` will be useful later. In the for loop at the end of this script,
    this df is called `q_scores`.
```{r}
get_all_hits <- function(query, tax_df) {
    n_tax <- nrow(tax_df)

  # Pre-allocate data frame rows for computational efficiency
  query_scores <- data.frame(
    min_dist = rep(NA_real_, n_tax),
    level = rep(NA_character_, n_tax),
    index = seq_len(n_tax),
    stringsAsFactors = FALSE
  )

  # Get metadata for all hits with the minimum distance score
  for (i in seq_len(n_tax)) {
    results <- get_row_results(query, tax_df, i)
    query_scores$min_dist[i] <- as.numeric(results[[1]])
    query_scores$level[i] <- results[[2]]
  }

  return(query_scores)
}
```

#### get_row_results
- Arguments:
  - `query`: A survey response, e.g. "neuroscience" or "AI".
  - `tax_df`: see get_all_hits
  - `row_num`: Row number of the row in `tax_df` you want to analyze.
- Details:
  - For a given query (survey response), and a given row in the taxonomy,
    record which column's entry had the lowest distance score. Breaks ties
    by just choosing the first occurrence of a minimum value. So it prefers
    Level1 over Level2, and Level2 over Level3. I'm betting that 99% of ties
    are basically just noise, and not real matches.
- Outputs:
  - A vector with two elements: first, the lowest distance score
    between the query and any string in that row of the taxonomy, and second,
    the name of the column that the winning string came from (e.g. `Level1`).
```{r}
get_row_results <- function(query, tax_df, row_num) {

  strL1 <- tax_df[row_num, "Level1"]
  strL2 <- tax_df[row_num, "Level2"]
  strL3 <- tax_df[row_num, "Level3"]

  # compute Levenshtein distances for each level
  dist1 <- as.numeric(
    adist(query, strL1)
  )
  dist2 <- as.numeric(
    ifelse(
      !is.na(strL2),
      adist(query, strL2),
      Inf
    )
  )
  dist3 <- as.numeric(
    ifelse(
      !is.na(strL3),
      adist(query, strL3),
      Inf
    )
  )
  
  dists <- c(dist1, dist2, dist3)
  min_dist <- min(dists, na.rm = TRUE)

  # which.min gets the index of the lowest element.
  # In the event of a tie, it chooses the first one.
  # So it will choose level1 over level2, and level2 over level3.
  level <- c("Level1", "Level2", "Level3")[
    which.min(dists)
  ]

  # This was a little test I ran to see how common ties are.
  # Turns out ties are extremely common. They are all, or nearly
  # all, cases where all 3 levels are really different from the query.
  # I'm willing to assume that a tie means the taxonomy strings are all
  # really different from the query, and I'm not going to fret over the
  # fact that this function is practically breaking the tie at random,
  # just choosing the first minimum in the list.
#   has_tied_min <- function(x, na.rm = FALSE) {
#   m <- min(x, na.rm = na.rm)
#   sum(x == m, na.rm = na.rm) > 1L
# }
  #if (has_tied_min(dists)) {message(query, " | ", tax_df[row_num,], " | ", dists)}

  return(
    c(min_dist, level)
  )
}
```

#### get_candidates
- Arguments:
  - `q_scores`: A data frame with columns `min_dist`, `level`, and `index`,
    produced by `get_all_hits()`.
- Details:
  - Given the minimum distance scores for every row/query pair, find the
    global minimum across all rows. Handles ties by reporting multiple
    candidates.
- Outputs:
  - A data frame with the same columns as the input data frame. It has been
    merely been subsetted so the only rows remaining are those with the
    minimum distance.
```{r}
get_candidates <- function(q_scores) {

  best_dist <- min(q_scores$min_dist)
  cands <- q_scores %>% filter(min_dist == best_dist)
  return(cands)
}
```

#### get_winner_df
- Arguments:
  - `cands_df`: A data frame with columns `min_dist`, `level`, and `index`,
    produced by `get_candidates()`.
  - `tax_df`: see `get_all_hits()`.
- Details:
  - Given a set of candidate rows, choose a smaller number of rows to keep.
    Chooses final picks by preferring Level 3 matches over Level 2 matches,
    and preferring Level 2 over Level 1. Returns a subset of the taxonomy,
    not just a bunch of indices and distance scores.
- Outputs:
  - A subset of the taxonomy; the final rows for manual inspection (as a
    data frame).
```{r}
get_winner_df <- function(cands_df, tax_df) {
  winners <- ""
  lvl3_cands <- subset(cands_df, level == "Level3")
  lvl2_cands <- subset(cands_df, level == "Level2")
  lvl1_cands <- subset(cands_df, level == "Level1")

  if (nrow(lvl3_cands) > 0) {
    winners <- tax_df[lvl3_cands$index, ]
  } else if (nrow(lvl2_cands) > 0) {
    keep_idx <- lvl2_cands$index[is.na(tax_df$Level3[lvl2_cands$index])]
    winners <- tax_df[keep_idx, , drop = FALSE]
  } else {
    keep_idx <- lvl1_cands$index[
      is.na(tax_df$Level2[lvl1_cands$index]) &
        is.na(tax_df$Level3[lvl1_cands$index])
    ]
    winners <- tax_df[keep_idx, , drop = FALSE]
  }
  return(winners)
}
```

# Load data
```{r}
other_quant <- load_qualtrics_data("clean_data/other_quant.tsv")
status <- load_qualtrics_data("clean_data/contributor_status_Q3.tsv")
qual <- load_qualtrics_data("qual_responses.tsv")
tax <- as.data.frame(
    readLines("data/digital_commons_disciplines.txt"),
    stringsAsFactors = FALSE)

data <- cbind(status, other_quant)
nrow(data)
head(data)
```

```{r}
tmp <- data$job_category[nzchar(data$job_category)]
job_count <- data.frame(table(tmp))
names(job_count) <- c("Job", "Count")

academics <- sum(subset(job_count, Job != "Non-research Staff")[, "Count"])
```

```{r}
qual_fields <- qual$subfield[nzchar(qual$subfield)]
length(qual_fields)
academics
```

This question was not mandatory, but 170/188 academics answered it.

```{r}
head(qual_fields)
```

Let's standardize the capitalization in this vector. The capitalization will now match my taxonomy, which is important since the fuzzy string matching algorithm I'll be using is case-sensitive.

```{r}
qual_fields <- unname(
    sapply(
        qual_fields,
        function(x) tools::toTitleCase(tolower(x)) )
    )

head(qual_fields)
```

Hmm. The correction of "AI" to "Ai" is unfortunate, but let's see how it goes.

Here's our taxonomy:

```{r}
head(tax)
```

Let's tidy this data frame.

```{r}
tax <- tax %>%
  separate(
    col   = names(tax)[1],
    into  = c("Level1", "Level2", "Level3"),
    sep   = ": ",
    fill  = "right",   # any missing pieces become NA
    extra = "merge"    # if there were >2 colons, they’d all merge into Level3
  )

head(tax)
```

You can take my word for it that I looked back and forth between the data frame and the data file, and made sure that the data frame looks good. (Correct # of rows, row numbers match line numbers, etc.)

Cool! Now we are ready for fuzzy string matching. Let's start with a very rudimentary method, and we can do something fancier if this appears insufficient.

We are just using the adist() function in base R to calculate the Levenshtein distance between pairs of strings--the minimum number of substitutions needed to turn 'string a' into 'string b'.

Since we invited people to separate multiple disciplines with a comma, we'll need to parse those and track participant IDs. Several people used a slash instead of a comma, so we'll use a regex that looks for that, too.
```{r}
# I think the data need to be a tibble for unnest to work

responses_df <- tibble(
  participantID = seq_along(qual_fields),
  response = strsplit(qual_fields, "\\s*[,/]\\s*", perl = TRUE)
  # \\s* matches zero or more whitespace (first backslash is an escape)
  # [,/] will match either a comma or a slash
  # With base R strsplit(), \\s only works if you set perl = TRUE
) %>%
  unnest_longer(response) %>% # takes a column of type list, where the entries
  # are themselves vectors, and turns each element of those vectors
  # into its own row, duplicating the other fields appropriately.
  mutate(response = trimws(response)) %>%     # strip whitespace around each field
  filter(response != "")                   # drop empty entries (from e.g. A,,B)

responses_df
```

Again, you can take my word for it that I visually compared this df to the raw survey data and it looks good. Now let's get to the meat of the script: for each discipline entered by a survey participant, get rows that contain the best possible fuzzy match (allowing multiple rows for ties).

This code takes a minute to run.
```{r}
results <- vector("list", nrow(responses_df))

for (i in seq_len(nrow(responses_df))) {
  current_query <- responses_df$response[i]
  current_participantID <- responses_df$participantID[i]

  q_scores <- get_all_hits(current_query, tax)
  candidates <- get_candidates(q_scores)
  winnerdf <- get_winner_df(candidates, tax)

  # add participant info
  winnerdf$participantID <- current_participantID
  winnerdf$response <- current_query
  results[[i]] <- winnerdf
}

final_results <- dplyr::bind_rows(results) %>%
  # move participantID to be the first column
  dplyr::relocate(participantID, response)

head(final_results, n=50)
```

Okay, well I got results, but there's a pretty high rate of crappy matches. And then there's some that only a sophisticated algorithm would catch, and a couple that I'M not even sure how to classify. I think experimenting with better algorithms, or "real ML" classifiers, would probably take longer than just manually sifting through these. My code is ugly, and I hate that my procedure is not reproducible, but it's fast. So I'm just going to manually review these classifications in Microsoft Excel.

```{r}
write.table(
  final_results,
  "data/qual_fields_guesses.tsv",
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)
```

```{r}
sessionInfo()
```
