---
title: "Future Contributors"
---

# Overview

Plotting results of Q15: "What would make you more likely to participate in OSS projects?". This question was only visible to respondents who said they haven't yet contributed to OS, but would like to do so in the future.

# Import packages and utilities

```{r}
project_root <- here::here() # requires that you be somewhere in the
# project directory (not above it)
# packages
suppressMessages(source(file.path(project_root, "scripts/packages.R")))
# functions and objects used across scripts
suppressMessages(source(file.path(project_root, "scripts/utils.R")))
```

# Load data
```{r}
future <- load_qualtrics_data("clean_data/future_contributors_Q15.tsv")
other_quant <- load_qualtrics_data("clean_data/other_quant.tsv")
head(future)
```

At this point, we COULD remove rows from participants who never saw this question, but since we're just tallying up the 1s, not the 0s, there's really no need.

# Prepare data for plotting
Sum up counts for each solution.
```{r}
to_plot <- data.frame(
  Solution = names(future),
  Count = unname(apply(future, 2, function(x) round(sum(x, na.rm = TRUE))))
)
to_plot
```

# Plot
For visual clarity, let's remove the "Other" row.
```{r}
to_plot <- to_plot %>% filter(Solution != "Other")
```

Reorder factor levels based on count.
```{r}
to_plot <- to_plot %>%
  mutate(Solution = fct_reorder(Solution, Count, .desc = FALSE))
```

And make a plot, using a function in utils.R.
```{r, fig.width=9, fig.height=6}
myplot <- basic_bar_chart(to_plot,
  x_var = "Solution",
  y_var = "Count",
  title = "What would help aspiring contributors get started?",
  horizontal = TRUE,
  show_bar_labels = TRUE,
  show_ticks_y = FALSE,
  color_index = 8,
  show_axis_title_x = TRUE,
  show_axis_title_y = FALSE,
  show_grid = TRUE
)
myplot
```

Save the plot if you wish.
```{r}
save_plot("future_contributors.tiff", 12, 8, p=myplot)
```

# Look at jobs of respondents
NOW let's remove rows where the participant never saw the question. Also, remove rows where we have no data on the participant's job.
```{r}
future_and_job <- future
future_and_job$job_category <- other_quant$job_category

n <- ncol(future)
future_cols <- names(future_and_job)[1:n]
last_col   <- names(future_and_job)[ncol(future_and_job)]

future_and_job <- future_and_job %>%
  filter(
    # drop rows where all of the future columns are 0
    !if_all(all_of(future_cols), ~ . == 0),
    # drop rows where the job column is ""
    .data[[last_col]] != ""
  )
```

Combine postdocs and other research staff for better visual clarity.
```{r}
future_and_job <- future_and_job %>%
  mutate(
    job_category = recode(
      job_category,
      "Post-Doc" = "Postdocs and\nStaff Researchers",
      "Other research staff" = "Postdocs and\nStaff Researchers"
    )
  )
```

At this point I'm not going to bother with additional wrangling because I'm just interested in the jobs of the people who answered this question. But we are poised to see which jobs voted for which solutions, if needed.
```{r}
jobs <- data.frame(table(future_and_job$job_category))
names(jobs) <- c("Job", "Count")
```

Reorder factor levels based on count.
```{r}
jobs <- jobs %>%
  mutate(Job = fct_reorder(Job, Count, .desc = TRUE))
```

And make a plot, using a function in utils.R.
```{r, fig.width=9, fig.height=6}
jobs_plot <- basic_bar_chart(jobs,
  x_var = "Job",
  y_var = "Count",
  title = "Composition of aspiring contributors by job",
  horizontal = FALSE,
  show_bar_labels = TRUE,
  show_ticks_y = FALSE,
  color_index = 2,
  show_axis_title_y = TRUE,
  show_grid = TRUE
)
jobs_plot
```

Save the plot if you wish.
```{r}
save_plot("future_contributors_jobs.tiff", 12, 8, p=jobs_plot)
```
