---
title: "Project sizes: exploratory plots"
---

# Overview
This notebook explores Q5: "How frequently have you contributed to projects of the following size?".

# Import packages and utilities
```{r}
project_root <- here::here() # requires that you be somewhere in the
# project directory (not above it)
# packages
suppressMessages(source(file.path(project_root, "scripts/packages.R")))
# functions and objects used across scripts
suppressMessages(source(file.path(project_root, "scripts/utils.R")))
```

# Load data
```{r}
sizes_raw <- load_qualtrics_data("clean_data/project_size_Q5.tsv")
other_quant <- load_qualtrics_data("clean_data/other_quant.tsv")
```

# Wrangle data

Drop rows with no data
```{r}
sizes <- exclude_empty_rows(sizes_raw)
nrow(sizes)
```

Let's create a long-format version for plotting.

```{r}
sizes_long <- sizes %>%
  pivot_longer(
    cols = everything(),
    names_to  = "size",
    values_to = "frequency"
  )

sizes_long
```

# Inspect data

Let's look at the counts.

```{r}
sizes_counts <- sizes_long %>%
  count(size, frequency, name = "n")

sizes_counts[
  order(
    sizes_counts$n,
    decreasing = TRUE
  ),
]
```


Reorder factor levels
```{r}
ordered_freqs <- c(
  "Never",
  "Relatively infrequently",
  "Occasionally",
  "Relatively frequently"
)

sizes_counts$frequency <- factor(
  sizes_counts$frequency,
  levels = ordered_freqs
)

ordered_sizes <- c(
  "Small",
  "Medium",
  "Large"
)

sizes_counts$size <- factor(
  sizes_counts$size,
  levels = ordered_sizes
)
```


# Stacked bar chart

```{r, fig.width=8, fig.height=8}
stacked_bar <- stacked_bar_chart(
  sizes_counts,
  x_var = "size",
  y_var = "n",
  fill = "frequency",
  title = "Relative Frequency of Contributions\nto Projects of a Certain Size",
  cpalette = c(
    "#1a7937", # dark green
    "#acd49f", # light green
    "#e5e5e5", # light gray
    "#c3a4d0", # light purple
    "#752a82" # dark purple
  )
)
stacked_bar <- stacked_bar +
  geom_hline(yintercept = 155.4, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 77.7, linetype = "dashed", color = "black")
stacked_bar
```

The dashed lines indicate 1/3 and 2/3 of the total number of responses.

Save the plot
```{r}
save_plot("proj_sizes_bar.tiff", 8, 6, p=stacked_bar)
```


# Incorporate job category

```{r}
sizes_job <- cbind(sizes_raw, other_quant$job_category)
# Rename column
names(sizes_job)[ncol(sizes_job)] <- "job_category"
# Filter out people who didn't answer either question
sizes_job <- exclude_empty_rows(sizes_job, strict = TRUE)
```

```{r}
sizes_job_long <- sizes_job %>%
  pivot_longer(
    cols = -job_category,
    names_to = "size",
    values_to = "frequency"
  )
```

```{r}
# three way cross tabs (xtabs) and flatten the table
# code from: https://ladal.edu.au/tutorials/regression/regression.html
ftable(xtabs(~ job_category + size + frequency, data = sizes_job_long))
```

For each job category, what percent of respondents in that category said they contribute to large projects occasionally or relatively frequently?

```{r}
high_freq <- c("Occasionally", "Relatively frequently")

pct_large_by_job <- sizes_job %>%
  group_by(job_category) %>%
  summarise(
    n_total   = n(),
    n_large   = sum(Large %in% high_freq, na.rm = TRUE),
    pct_large = round(100 * mean(Large %in% high_freq, na.rm = TRUE), 1)
  ) %>%
  arrange(desc(pct_large))

pct_large_by_job
```

Let's do the same for small projects.

```{r}
pct_small_by_job <- sizes_job %>%
  group_by(job_category) %>%
  summarise(
    n_total   = n(),
    n_small   = sum(Small %in% high_freq, na.rm = TRUE),
    pct_small = round(100 * mean(Small %in% high_freq, na.rm = TRUE), 1)
  ) %>%
  arrange(desc(pct_small))

pct_small_by_job
```

Merge the two data frames.

```{r}
merged <- inner_join(
  pct_small_by_job,
  pct_large_by_job,
  by = c("job_category", "n_total")
) %>%
  select(job_category, n_total, n_small, pct_small, n_large, pct_large)

merged
```

Save for supplement.

```{r}
write_df_to_file(merged, "supplementary_tables/project_sizes_perc_by_job.tsv")
```

What if we add a row for academics? (Just summing up all rows except nr staff)

```{r}
tmp <- pct_large_by_job %>%
  filter(job_category != "Non-research Staff")

rbind(
  tmp,
  c(
    "Academic",
    sum(tmp$n_total),
    sum(tmp$n_large),
    round(sum(tmp$n_large) / sum(tmp$n_total) * 100, 2)
  )
)

```

Maybe we should fold in the smaller job categories, like we did with the regressions.

```{r}
combined <- sizes_job_long %>%
  mutate(
    job_category = recode(
      job_category,
      "Post-Doc" = "Postdocs and Staff Researchers",
      "Other research staff" = "Postdocs and Staff Researchers"
    )
  )

combined <- combined %>%
  mutate(
    job_category = recode(
      job_category,
      "Grad Student" = "Students",
      "Undergraduate" = "Students"
    )
  )
```

What if we separated this stacked bar into academics vs. non-research staff (or IT, maybe)? Maybe just do small and large, to make things visually simpler. Let's build each plot separately and then stitch them together using patchwork.

```{r}
# A version of the df where all academics
# have been relabeled to academic
combined_acad_nrstaff <- combined %>%
  mutate(
    job_category = recode(
      job_category,
      "Students" = "Academic",
      "Postdocs and Staff Researchers" = "Academic",
      "Faculty" = "Academic"
    )
  )

combined_acad_nrstaff$frequency <- factor(
  combined_acad_nrstaff$frequency, 
  levels = ordered_freqs
  )

acad_counts <- combined_acad_nrstaff %>%
  filter(job_category == "Academic") %>%
  filter(size != "Medium") %>%
  count(size, frequency, name = "n")

nrstaff_counts <- combined_acad_nrstaff %>%
  filter(job_category != "Academic") %>%
  filter(size != "Medium") %>%
  count(size, frequency, name = "n")
```

Save data frames for fine-tuning of figure in a separate script.
```{r}
write_df_to_file(acad_counts, "data_for_plots/project_sizes_acad.tsv")
write_df_to_file(nrstaff_counts, "data_for_plots/project_sizes_staff.tsv")
```

```{r, fig.width=8, fig.height=8}
stacked_bar_acad <- stacked_bar_chart(
  acad_counts,
  x_var = "size",
  y_var = "n",
  fill = "frequency",
  title = "Academics",
  ylabel = "Percent of Responses",
  proportional = TRUE,
  show_legend = FALSE,
  cpalette = c(
    "#1a7937", # dark green
    "#acd49f", # light green
    "#e5e5e5", # light gray
    "#c3a4d0", # light purple
    "#752a82" # dark purple
  )
)

stacked_bar_acad <- stacked_bar_acad +
  scale_y_continuous(labels = scales::percent)
# stacked_bar_acad <- stacked_bar_acad +
#   geom_hline(yintercept = 155.4, linetype = "dashed", color = "black") +
#   geom_hline(yintercept = 77.7, linetype = "dashed", color = "black")
stacked_bar_acad
```


```{r, fig.width=8, fig.height=8}
stacked_bar_nrstaff <- stacked_bar_chart(
  nrstaff_counts,
  x_var = "size",
  y_var = "n",
  fill = "frequency",
  title = "Non-research Staff",
  ylabel = "Percent of Responses",
  proportional = TRUE,
  show_axis_title_y = FALSE,
  cpalette = c(
    "#1a7937", # dark green
    "#acd49f", # light green
    "#e5e5e5", # light gray
    "#c3a4d0", # light purple
    "#752a82" # dark purple
  )
)

stacked_bar_nrstaff <- stacked_bar_nrstaff +
  scale_y_continuous(labels = scales::percent)

stacked_bar_nrstaff
```

Combine onto one plot

```{r, fig.width=10, fig.height=6}
p_combined <- patchwork::wrap_plots(
  stacked_bar_acad,
  stacked_bar_nrstaff
) +
  plot_annotation(
    title = "Frequency of Contributions to Projects of a Certain Size",
    theme = theme(plot.title = element_text(size = 24, face = "bold"))
  )

p_combined
```

Save the plot
```{r}
save_plot("proj_sizes_acad_nrstaff.tiff", 14, 6, p=p_combined)
```

What if we include IT? Start with non-research staff.

```{r}
sizes_staff <- cbind(sizes_raw, other_quant$staff_categories)
# Rename column
names(sizes_staff)[ncol(sizes_staff)] <- "staff_categories"
# Filter out people who didn't answer either question
sizes_staff <- exclude_empty_rows(sizes_staff, strict = TRUE)

nrow(sizes_staff)
head(sizes_staff)
```

Now limit to just IT.

```{r}
sizes_it <- sizes_staff %>%
  filter(staff_categories == "Information Technology (IT)")

nrow(sizes_it)
```

```{r}
sizes_it_long <- sizes_it %>%
  pivot_longer(
    cols = -staff_categories,
    names_to = "size",
    values_to = "frequency"
  ) %>%
    select(-staff_categories)

sizes_it_long
```

```{r}
it_counts <- sizes_it_long %>%
  filter(size != "Medium") %>%
  count(size, frequency, name = "n")

# Re-order factor levels
#it_counts$size <- factor(it_counts$size, levels = c("Small", "Large"))
it_counts$frequency <- factor(it_counts$frequency, levels = ordered_freqs)
it_counts
```


```{r, fig.width=8, fig.height=8}
stacked_bar_it <- stacked_bar_chart(
  it_counts,
  x_var = "size",
  y_var = "n",
  fill = "frequency",
  title = "IT Staff",
  ylabel = "Percent of Responses",
  proportional = TRUE,
  show_axis_title_y = FALSE,
  cpalette = c(
    "#1a7937", # dark green
    "#acd49f", # light green
    "#e5e5e5", # light gray
    "#c3a4d0", # light purple
    "#752a82" # dark purple
  )
)

stacked_bar_it <- stacked_bar_it +
  scale_y_continuous(labels = scales::percent)

stacked_bar_it
```

Meh, not that interesting. Looks very similar to the plot for non-research staff. Save it anyway.

Save the plot
```{r}
save_plot("proj_sizes_it.tiff", 14, 6, p=stacked_bar_it)
```




# Line plots

Reorder factor levels
```{r}
ordered_jobs <- c(
  "Students",
  "Postdocs and Staff Researchers",
  "Faculty",
  "Non-research Staff"
)

combined$size <- factor(combined$size, levels = ordered_sizes)
combined$frequency <- factor(combined$frequency, levels = ordered_freqs)
combined$job_category <- factor(combined$job_category, levels = ordered_jobs)
```

Recode frequency from categorical to a numeric score

```{r}
combined_coded_all <- combined %>%
  mutate(
    frequency_score = recode(
      frequency,
      "Never" = 0L,
      "Relatively infrequently" = 1L,
      "Occasionally" = 2L,
      "Relatively frequently" = 3L
    )
  ) %>%
    select(-frequency)

combined_coded_all
```

Sum up frequency scores

```{r}
combined_scores <- combined_coded_all %>%
  count(job_category, size, wt = frequency_score, name = "total_score")

# Reorder factor levels
combined_scores$size <- factor(combined_scores$size, levels = ordered_sizes)

combined_scores
```


```{r, fig.width=8, fig.height=8}
ggplot(
    combined_scores,
    aes(x = size, y = total_score, group = job_category, color = job_category)
  ) +
    geom_line() +
    geom_point() +
    ylim(0, 175) +
    scale_x_discrete(expand = c(0.025, 0.025)) +
    ylab("Frequency Score") +
    xlab("Project Size") +
    ggtitle("Frequency of Contributions by Project Size") +
    scale_color_manual(values = COLORS) +
    
    theme(
          axis.title.y = element_text(
            angle = 0,
            vjust = 0.5,
            size = 18,
            face = "bold"
          ),
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.background = element_blank(),
          panel.grid.major = element_line(linewidth = 0.25, linetype = "solid", color = "gray90"),
          panel.grid.minor = element_line(linewidth = 0.25, linetype = "solid", color = "gray90"),
          plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
          plot.title = element_text(hjust = 0.5, size = 16),
        )
```

Nah, still needs work. How about we just plot the trend for large projects?

# Large projects

```{r}
large <- subset(combined, size == "Large")
large_counts <- large %>%
  count(job_category, frequency, name = "n")

large_counts <- large_counts %>%
  group_by(job_category) %>%
  mutate(perc_total = round(100 * n / sum(n), 1)) %>%
  ungroup()
```


```{r, fig.width=8, fig.height=8}
large_line <- ggplot(
    large_counts,
    aes(x = frequency, y = perc_total, group = job_category, color = job_category)
  ) +
    geom_line() +
    geom_point() +
    ylim(0, 50) +

    scale_x_discrete(expand = c(0.025, 0.025)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1, scale = 1), limits = c(0, 60)) +
    scale_color_manual(values = COLORS) +

    ylab("Percent of Respondents in Job Category") +
    xlab("Project Size") +
    ggtitle("Frequency of Contributions\nto Large Projects") +
    
    
    theme(
          axis.title.y = element_text(
            size = 22,
            margin = margin(r = 20)
          ),
          axis.text.y = element_text(size = 20),
          axis.title.x = element_blank(),
          axis.text.x = element_text(
            angle = -45,
            hjust = 0,
            vjust = 1,
            size = 20,
            margin = margin(t = 6)),
          #axis.ticks.x = element_blank(),
          legend.text = element_text(size = 20),
          legend.title = element_blank(),
          panel.background = element_blank(),
          panel.grid.major = element_line(linewidth = 0.25, linetype = "solid", color = "gray90"),
          panel.grid.minor = element_line(linewidth = 0.25, linetype = "solid", color = "gray90"),
          plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
          plot.title = element_text(hjust = 0, size = 24, face = "bold"),
        )

large_line
```

Hard to discern a clear trend. Let's save the plot anyway.

Save the plot
```{r}
save_plot("proj_sizes_large_line.tiff", 10, 6, p=large_line)
```


# Medium projects
What about Medium projects? Do the same trends hold?
```{r}
med <- subset(combined, size == "Medium")
med_counts <- med %>%
  count(job_category, frequency, name = "n")

med_counts <- med_counts %>%
  group_by(job_category) %>%
  mutate(perc_total = round(100 * n / sum(n), 1)) %>%
  ungroup()
```

```{r, fig.width=8, fig.height=8}
med_line <- ggplot(
    med_counts,
    aes(x = frequency, y = perc_total, group = job_category, color = job_category)
  ) +
    geom_line() +
    geom_point() +
    ylim(0, 50) +

    scale_x_discrete(expand = c(0.025, 0.025)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1, scale = 1), limits = c(0, 60)) +
    scale_color_manual(values = COLORS) +

    ylab("Percent of Respondents in Job Category") +
    xlab("Project Size") +
    ggtitle("Frequency of Contributions\nto Medium Projects") +
    
    
    theme(
          axis.title.y = element_text(
            size = 22,
            margin = margin(r = 20)
          ),
          axis.text.y = element_text(size = 20),
          axis.title.x = element_blank(),
          axis.text.x = element_text(
            angle = -45,
            hjust = 0,
            vjust = 1,
            size = 20,
            margin = margin(t = 6)),
          #axis.ticks.x = element_blank(),
          legend.text = element_text(size = 20),
          legend.title = element_blank(),
          panel.background = element_blank(),
          panel.grid.major = element_line(linewidth = 0.25, linetype = "solid", color = "gray90"),
          panel.grid.minor = element_line(linewidth = 0.25, linetype = "solid", color = "gray90"),
          plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
          plot.title = element_text(hjust = 0, size = 24, face = "bold"),
        )

med_line
```

Save the plot
```{r}
save_plot("proj_sizes_med_line.tiff", 10, 6, p=med_line)
```

# Small projects
We've made it this far. We might as well look at small projects, too.

```{r}
small <- subset(combined, size == "Small")
small_counts <- small %>%
  count(job_category, frequency, name = "n")

small_counts <- small_counts %>%
  group_by(job_category) %>%
  mutate(perc_total = round(100 * n / sum(n), 1)) %>%
  ungroup()
```

```{r, fig.width=8, fig.height=8}
small_line <- ggplot(
    small_counts,
    aes(x = frequency, y = perc_total, group = job_category, color = job_category)
  ) +
    geom_line() +
    geom_point() +
    ylim(0, 50) +

    scale_x_discrete(expand = c(0.025, 0.025)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1, scale = 1), limits = c(0, 60)) +
    scale_color_manual(values = COLORS) +

    ylab("Percent of Respondents in Job Category") +
    xlab("Project Size") +
    ggtitle("Frequency of Contributions\nto Small Projects") +
    
    
    theme(
          axis.title.y = element_text(
            size = 22,
            margin = margin(r = 20)
          ),
          axis.text.y = element_text(size = 20),
          axis.title.x = element_blank(),
          axis.text.x = element_text(
            angle = -45,
            hjust = 0,
            vjust = 1,
            size = 20,
            margin = margin(t = 6)),
          #axis.ticks.x = element_blank(),
          legend.text = element_text(size = 20),
          legend.title = element_blank(),
          panel.background = element_blank(),
          panel.grid.major = element_line(linewidth = 0.25, linetype = "solid", color = "gray90"),
          panel.grid.minor = element_line(linewidth = 0.25, linetype = "solid", color = "gray90"),
          plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
          plot.title = element_text(hjust = 0, size = 24, face = "bold"),
        )

small_line
```

Wow, that's much prettier.

Save the plot
```{r}
save_plot("proj_sizes_small_line.tiff", 10, 6, p=small_line)
```


```{r}
p_combined <- patchwork::wrap_plots(large_line, plot_spacer(), small_line) +
  plot_layout(widths = c(1, 0.05, 1))

p_combined <- p_combined +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 26))

svglite::svglite(
  file.path(FIGURE_PATH, "figureS3.svg"),
  width = 26,
  height = 10
)
print(p_combined)
dev.off()
```


I'd like to know whether the proportion of academics who contribute to large projects with some frequency is significantly lower than the proportion of non-research staff who contribute to large projects with some frequency.

```{r}
combined_counts <- combined %>%
  count(job_category, size, frequency, name = "n")

res <- combined_counts %>%
  filter(size == "Large") %>%
  mutate(
    group = if_else(job_category == "Non-research Staff",
                    "Non-research Staff", "Academics"),
    freq2 = if_else(frequency == "Never", "Never", "Other")
  ) %>%
  group_by(group, freq2) %>%
  summarise(n = sum(n), .groups = "drop_last") %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

# 2×2 table: proportions for each group
res_wide <- res %>%
  select(group, freq2, prop) %>%
  pivot_wider(names_from = freq2, values_from = prop) %>%
  arrange(match(group, c("Non-research Staff", "Academics")))

res_wide
```

Hmm. Seems promising. We should probably do a regression...
